---
title: "Gjennomføring etter kjønn"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
#options(scipen = 2)
options(digits = 2)
options(scipen=999)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggbeeswarm)

```


## Functions

local functions:
  specify_decimal()
  %notin%
  select_prikk()
  map_levels()
  minmax()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```


C:/Users/HANN/OneDrive - Helsedirektoratet/
O:/Prosjekt/FHP/

GJENNOMFORING_VGS_INNVKAT_2024-09-18-10-34.csv

## Data

GEO
Jointable
recent year
```{r}
stednavn_GEO_BKFL <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_BYDEL.csv", delim = ";")

```



KUBER

```{r}
file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/GJENNOMFORING_VGS_INNVKAT_3_2025-11-17-18-54.csv"
#file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/KOMMUNEHELSA/DATERT/csv/GJENNOMFORING_VGS_INNVKAT_3_2025-01-31-21-10.csv"

gjennomforing <- read_delim(file, delim = ";", locale = locale(encoding = 'ISO-8859-15')) %>% 
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL)

gjennomforing_data <- gsub(".csv", "", basename(file))

```

FHI colors
```{r}
fhiblue3 <- "#393c61"
fhiblue2 <- "#0975B5"
fhiblue2b <- "#0580A7"
fhiblue1 <- "#2EA1C0"
fhiblue1b <- "#65ADC3"
fhigrey <- "#A9B7BF"
fhigreen <- "#709900"
fhidarkgreen <- "#4d7e13"
fhilightgreen <- "#aed267"
fhilightgreenman <- "#a2c24c"
fhired <- "#B11643"
fhicoral <- "#FC5F56"
fhiyellow <- "#F7B500"


show_col(fhilightgreenman)

show_col(c(fhiblue1, fhiblue2, fhiblue3,
           fhiblue1b, fhiblue2b, fhigrey,
           fhilightgreenman, fhigreen, fhidarkgreen,
           fhired, fhicoral, fhiyellow),
         ncol = 3)

colors_LKB <- c(fhiblue3, fhiblue2b, fhilightgreen)
show_col(colors_LKB, ncol = 3)
```


Organisert fritidsaktivitet, etter sosioøkonomisk status

# Data prep

filtering

```{r}
gjennomforing %>% 
  map_levels()
```

```{r}
gjennomforing_prep <- gjennomforing %>% 
  filter(AAR_end == max(AAR_end),
         INNVKAT == 0) %>% 
  mutate(KJONN_tekst = case_when(KJONN == 0 ~ "Kjønn samlet",
                                 KJONN == 1 ~ "Gutter",
                                 KJONN == 2 ~ "Jenter",
                                 ))

```

```{r}
gjennomforing_plot <- left_join(jointable, gjennomforing_prep) %>% 
  filter(GEO_type %in% c("land", "bydel", "kommune")) %>% 
  mutate(prikk = ifelse(is.na(MEIS), 3, NA_real_)) %>% 
  mutate(group = interaction(reorder(Sted, plotorder), reorder(KJONN_tekst, KJONN), sep = "\n"),
         KJONN_group = case_when(KJONN %in% c(1,2) ~ 0,
                                 KJONN == 0 ~ 1),
         ) %>% 
  mutate(plotid = ifelse(grepl("^301", plotid), paste0("0", plotid), plotid),
         Sted = case_when(GEO == 301 ~ paste0(Sted, " (kommune)"),
                          GEO == 3 ~ paste0(Sted, " (fylke)"),
                          TRUE ~ Sted)) # ensure 4 digit Oslo geo code

gjennomforing_plot
```

missing
```{r}
gjennomforing_prep %>% 
  filter(GEO_type == "bydel") %>% 
  group_by(KJONN) %>% 
  count(SPVFLAGG)


select_prikk(gjennomforing_plot, GEO_type = "bydel")
```

# PLOT

## Test kommune - facet blue/green


facet by KJONN group (total and subgroupes)

```{r fig.height=5, fig.width=7}

x_text_size <- 13.5
y_text_size <- 12
axis_title_size <- 14
legend_size <- 14
ymax <- 100
ylab <- "Prosent gjennomført VGO, standardisert"
kilde <- "Kilde: SSB"
kildeaar <- gjennomforing_prep %>% 
  mutate(kildeaar = paste0(AAR_start, "-", AAR_end)) %>% 
  select(kildeaar) %>% 
  unique() %>% 
  pull
kildelab <- paste0(kilde, " (", kildeaar, ")")
  
#colors_LFK <- c(fhiblue2, fhiblue3, fhigreen)

theme_OVP <- theme_cowplot() +
    theme(legend.position = "bottom",
          legend.text = element_text(size = legend_size),
          legend.margin=margin(t=-6),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = x_text_size,
                                     #angle = 45, hjust = 1,
                                     ),
          axis.text.y = element_text(size = y_text_size),
          axis.title.y = element_text(size = axis_title_size),
          strip.background = element_blank(),
          #strip.text = element_blank(),
          line = element_blank(),
          axis.ticks.y = element_line(color = "grey50"),
          panel.border = element_rect(fill = NA, color = "grey50"),
          plot.caption = element_text(hjust=c(0,1), size = c(5,10), color = c("grey50", "black")),
          plot.margin = margin(t = 2, r = 5, b = 2, l = 5),
          panel.background = element_rect(color = "white"), 
          )


#misstext <- "Manglende eller            \n utilstrekkelig tallgrunnlag"
#misstext <- "Manglende eller anonymiserte tall"
misstext <- "Manglende eller for lite tallgrunnlag"

do_plot <- function(data, include_missing = FALSE) {  # default is missing FALSE, no need to speficy
  
    highest_plotorder <- data %>% filter(plotorder == max(plotorder)) %>% pull(Sted)
  
    p <- data %>% 
    ggplot(aes(x = reorder(KJONN_tekst, KJONN), 
               y = MEIS,
               fill = reorder(Sted,plotorder),
               
               )) +
    geom_col(aes(label = round(MEIS, 0),
               ),
             width = 0.6,
             position = position_dodge2(preserve = "single")) +
    scale_fill_manual(values = colors_LKB) +
    #scale_alpha_manual(values = c(0.5,1)) +
    geom_text(aes(label = round(MEIS, 0)),
              vjust = -0.6,
              position = position_dodge(0.6),
              size = 4.2,
              ) +
    scale_x_discrete(expand = expansion(mult = c(0.5, 0.5))) +
    scale_y_continuous(breaks = seq(0,100,10),
                       limits = c(0,ymax+1), expand = expansion(mult = c(0, 0))) +
    # coord_cartesian(ylim = c(0,100),
    #                 #xlim = c(1,3),
    #                 clip = "off") +
  theme_OVP +
    theme(strip.text = element_blank()) +
    labs(y = ylab,
         caption = c(gjennomforing_data, kildelab),
         fill = "") +
      facet_wrap(~KJONN_group, scales = "free_x", space = "free_x")
    
    
  if (include_missing) {
    
    facet_missing <- data %>% 
      filter(KJONN_group == 0,
             !is.na(prikk))
    
    p <- p +
      geom_point(aes(y = prikk,
                     group = interaction(reorder(KJONN_tekst, KJONN), reorder(Sted,plotorder)),
                     ), 
                 color = "black", size = 4.1,
                 position = position_dodge(width = 0.6),
                 show.legend = FALSE) +
       geom_text(data = facet_missing,
              aes(label = misstext,
                  ),
                 x = 1.25,
                 y = ymax-4,
                 size = 4,
                 check_overlap = TRUE,
                 inherit.aes = FALSE) +
       geom_point(data = facet_missing,
              #aes(group = group),
                  x = 0.60, y = ymax-4,
                  size = 4,
                  inherit.aes = FALSE)
  }
  
  return(p)
}

gjennomforing_plot %>% 
  filter(plotid == 110302) %>% 
  do_plot(., include_missing = FALSE)
ggsave("plots/Figur4_gjennomføring_kjønn_bydel.png", height = 5, width = 7, bg = "white")

gjennomforing_plot %>% 
  filter(plotid == 110308) %>% 
  do_plot(., include_missing = TRUE)
ggsave("plots/Figur4_gjennomføring_kjønn_bydel_missing.png", height = 5, width = 7, bg = "white")

```


## Missing

```{r}
missing <- select_prikk(gjennomforing_plot, GEO_type = "bydel")

length(missing)
missing
```


# MULTIPLOT


```{r}
# Figure directory name and figure position
dirname <- "Figure4_BYDEL"
figposnr <- "5"


# Generate plots without missing values
plots <- gjennomforing_plot %>%
  filter(!plotid %in% missing) %>%  # Exclude missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., include_missing = FALSE))  # Normal plot

# Generate plots with missing values
plots_prikk <- gjennomforing_plot %>%
  filter(plotid %in% missing) %>%  # Only include missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., include_missing = TRUE))  # Include missing-value markers

# Combine both lists of plots
all_plots <- c(plots, plots_prikk)

# Generate file paths using `outdir` and `figposnr`
paths <- stringr::str_c(names(all_plots), "_", figposnr, "_tema", ".png")

# Create the output directory based on `outdir`
res_dir <- paste0("plots/", dirname)
dir.create(res_dir, recursive = TRUE)

# Save plots
pwalk(list(paths, all_plots), ~ggsave(filename = .x, plot = .y, 
                                      height = 5, width = 7, dpi = 200, bg = "white", 
                                      path = res_dir))

length(all_plots)
```

