---
title: "Ungdatafigur 2026"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
#options(scipen = 2)
options(digits = 2)
options(scipen=999)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggrepel)

```

O:/Prosjekt/FHP/
C:/Users/HANN/OneDrive - Helsedirektoratet/

## Functions

local functions:
  specify_decimal()
  %notint%
  select_prikk()
  map_levels()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```


## GEO  
recent year ungdata
```{r}
stednavn_GEO_BKFL_GEO2024 <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_GEO2024.csv", delim = ";")

recentyear_KH2026_fylke_by <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/recentyear_fylke_KH2026.csv", delim = ";")

recentyear_KH2026 <-  read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/recentyear_KH2026_KB.csv", delim = ";")
```

```{r}
recentyear_KH2024 %>% 
  group_by(AAR_end, completed) %>% 
  count()
```

```{r}
jointable
```

```{r}
recentyear_KH2026
```


## Data
ungdata indikatorer
O:/Prosjekt/FHP/
O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2025
```{r}
# select the right data
#data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2025/"  # path to the data
#files_1 <- dir(data_path, pattern = ".*UNGD.*.csv") # get file names

data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/"  # path to the data
files <- dir(data_path, pattern = ".*UNGDATA.*.csv") # get file names

ungdata <- data_frame(filename = files) %>% # create a data frame
                                         # holding the file names
  mutate(file_contents = map(filename,          # read files into
           ~ read_delim(file.path(data_path, .), delim = ";")) # a new data column
        )  %>% 
  unnest() %>% 
  separate(filename, into = c("filgruppe"), sep = "_202", extra = "drop", remove = TRUE) %>%
  mutate(filgruppe = gsub("_UNGD.*", "", filgruppe)) %>%
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL_GEO2024)

```

```{r}
map_levels_bycols(ungdata, bycols = "filgruppe")
```
```{r}
ungdata %>% 
  count(filgruppe, KJONN, SOES)
```



# colors

FHI colors
```{r}
fhiblue3 <- "#393c61"
fhiblue2 <- "#0975B5"
fhiblue2b <- "#0580A7"
fhiblue1 <- "#2EA1C0"
fhiblue1b <- "#65ADC3"
fhigrey <- "#A9B7BF"
fhigreen <- "#709900"
fhidarkgreen <- "#4d7e13"
fhilightgreen <- "#aed267"
fhilightgreenman <- "#a2c24c"
fhired <- "#B11643"
fhicoral <- "#FC5F56"
fhiyellow <- "#F7B500"

show_col(c(fhiblue1, fhiblue2, fhiblue3,
           fhigrey, fhigreen, fhiblue3,
           fhired, fhicoral, fhiyellow),
         ncol = 3)

ungcolors <- c(fhigreen, fhiblue2, fhiblue1b, fhiblue3, fhiyellow, fhired)
show_col(ungcolors)
```

# Ungdata prep

```{r}
names(ungdata)
```

```{r}
map_levels(ungdata)
```


## Select indikators and filtering

```{r}
ungdata_prep <- ungdata %>% 
  filter(grepl("ALKO|HASJBRUK|NARKO", filgruppe),
         SOES == 0,
         ) %>% 
  
  mutate(kjonn_text = case_when(KJONN == 0 ~ "Total",
                                KJONN == 1 ~ "Gutter",
                                KJONN == 2 ~ "Jenter"),
         soes_text = case_when(SOES == 0 ~ "Total",
                               SOES == 1 ~ "Lav",
                               SOES == 2 ~ "Middels",
                               SOES == 3 ~ "Høy"))

```


# Overview plot hele landet + kjønn

```{r}
min_max <- minmax(ungdata_prep)
min_max
```


```{r fig.height=5, fig.width=7}
ungdata_prep %>% 
  filter(GEO == 0,
         SOES %notin% c(1:3)) %>% 
  ggplot(aes(x = AAR_end,
             y = MEIS,
             color = kjonn_text,
             group = interaction(kjonn_text, filgruppe)),
             ) +
  geom_line() +
  scale_x_continuous(breaks = seq(min_max$min_aar, min_max$max_aar, 2)) +
  scale_color_manual(values = c("Total" = fhigrey, "Gutter" = fhiblue2, "Jenter" = fhired)) +
  theme(legend.position = "bottom") +
  facet_wrap(~filgruppe, scales = "free_y") +
  labs(title = "Ungdataindikatorer - utvikling over tid",
       subtitle = "Landsgjennomsnitt",
       y = "Andel(%), standardisert", 
       x = "År")
ggsave("plots/Rusindikatorer_overtid_kjonn.png", height = 6, width = 9)
```


## PLOT TEST

## alkohol, cannabis, andre narko

```{r}
ungdata_recentyear_plot <- left_join(jointable, ungdata_prep) %>% 
  filter(GEO_type %in% c("land", "fylke", "kommune")) %>% 
    right_join(recentyear_KH2026 %>% 
                 filter(GEO_type == "kommune") %>% 
               transmute(plotid = GEO, AAR_end)) %>% 
  mutate(prikk = ifelse(is.na(MEIS), 1.5, NA_real_)) %>% 
  mutate(label_ungdata = paste0("Ungdata ", AAR_end),
         label_kilde = paste0("Kilde: Ungdata (", AAR_end, ")")) %>% 
  mutate(plotid = ifelse(plotid == 301, "0301", plotid),                # ensure 4 digit Oslo geo code
         Sted = case_when(GEO == 301 ~ paste0(Sted, " (kommune)"),      # separate Oslo fylke and kommune
                          GEO == 3 ~ paste0(Sted, " (fylke)"),
                          TRUE ~ Sted)) %>% 
  mutate(Sted_plotfylke = ifelse(GEO_type == "fylke", 
                                 paste0(Sted, " (kommunetall mangler, fylke vises)"), 
                                 Sted)) %>%
  mutate(varorder = case_when(grepl("ROYK", filgruppe) ~ 1,
                             grepl("SNUS", filgruppe) ~ 2,
                             grepl("VAPE", filgruppe) ~ 3,
                             grepl("ALKOHOL", filgruppe) ~ 4,
                             grepl("HASJBRUK", filgruppe) ~ 5,
                             grepl("ANDRENARKO", filgruppe) ~ 6,
                             TRUE ~ NA_real_
                             ),
        vartext = case_when(grepl("ROYK", filgruppe) ~ "Røyker \nukentlig/daglig",
                               grepl("SNUS", filgruppe) ~ "Snuser \nukentlig/daglig",
                               grepl("VAPE", filgruppe) ~ "Vaper \nukentlig/daglig",
                               grepl("ALKOHOL", filgruppe) ~ "Har vært \nberuset på alkohol",
                               grepl("HASJBRUK", filgruppe) ~ "Har prøvd \nhasj/cannabis",
                               grepl("ANDRENARKO", filgruppe) ~ "Har prøvd andre \nnarkotiske stoff",
                               )) %>% 
  mutate(plotid = ifelse(plotid == 301, "0301", plotid))

```

```{r fig.height=5, fig.width=7}
ylab <- "Prosent ungdomsskoleelever, standardisert"

plotid_to_plot <- 5001

label_kilde <- ungdata_recentyear_plot %>% 
  filter(plotid == 5001) %>% 
  pull(label_kilde)

ungdata_recentyear_plot %>% 
  filter(plotid == 5001,
         GEO_type %in% c("land", "kommune")) %>%
  ggplot(aes(x = kjonn_text, y = MEIS, fill = Sted)) +
  geom_col(position = position_dodge(),
           width = 0.7) +
  geom_text(aes(label = round(MEIS, 0)),
              position = position_dodge(width = 0.7),
              vjust = -1,
              size = 4.2,
              ) +
  scale_fill_manual(values = c(fhiblue3, fhiblue1b)) +
  #scale_fill_manual(values = c("Total" = fhigrey, "Gutter" = fhiblue2, "Jenter" = fhired)) +
  scale_y_continuous(limits = c(0,20)) +
  facet_wrap(~vartext) +
  labs(y = ylab,
       x = "",
       fill = "",
       caption = label_kilde[1],
       ) +
  theme_cowplot() +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
        legend.position = "bottom",
        axis.title.x = element_blank(),
        )
ggsave("plots/rus_landet_kjonn_sted.png", height = 5, width = 7, bg = "white")
```

# MULTIPLOT

```{r}
ungdata_recentyear_plot %>% 
  count(filgruppe)
```

```{r}
metadata <- sub(pattern = ".csv", "", basename(files)[grepl("ALKOHOL|ANDRENARKO|HASJBRUK", basename(files))])

label_metadata <- paste0(metadata, collapse = ", ")
label_metadata

```
```{r}
min_max <- minmax(ungdata_recentyear_plot)
min_max

ungdata_recentyear_plot %>% 
  group_by(filgruppe) %>% 
  minmax(.)
ungdata_recentyear_plot %>% 
  arrange(-MEIS)
```

```{r fig.height=5, fig.width=7}

x_text_size <- 12.5
y_text_size <- 12
axis_title_size <- 14
legend_size <- 14
stripsize <- 14
dodgewidth <- 0.75
ymax <- min_max$max_meis+3
ylab <- "Prosent ungdomsskoleelever, standardisert"

theme_OVP <- theme_cowplot() +
    theme(legend.position = "bottom",
          legend.text = element_text(size = legend_size),
          legend.margin=margin(t=-6),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = x_text_size,
                                     #angle = 45, hjust = 1,
                                     ),
          axis.text.y = element_text(size = y_text_size),
          axis.title.y = element_text(size = axis_title_size),
          strip.background = element_blank(),
          strip.text = element_text(size = stripsize),
          line = element_blank(),
          axis.ticks.y = element_line(color = "grey50"),
          panel.border = element_rect(fill = NA, color = "grey50"),
          plot.caption = element_text(hjust=c(0,1), size = c(5,10), color = c("grey", "black")),
          plot.margin = margin(t = 2, r = 5, b = 2, l = 5),
          panel.background = element_rect(color = "white"), 
          )

#misstext <- "Manglende eller \nanonymiserte tall"
misstext <- "Manglende eller    \nfor lite tallgrunnlag"

do_plot <- function(data, include_missing = FALSE, missing_kommune = FALSE) {
  
  # Velg geografi og variabel til fill
  geo_keep <- if (missing_kommune) c("land", "fylke") else c("land", "kommune")
  fill_col <- if (missing_kommune) "Sted_plotfylke" else "Sted"
  # Bruk "Sted_plotfylke" her hvis det er riktig kolonne:
  # fill_col <- if (missing_kommune) "Sted_plotfylke" else "Sted"
  
  # Filtert datasett
  d <- data %>%
    filter(GEO_type %in% geo_keep)
  
  # Hent label_kilde
  label_kilde <- d %>% pull(label_kilde) %>% unique()
  
  # Grunnplott
  p <- d %>%
    ggplot(aes(
      x = kjonn_text,
      y = MEIS,
      fill = reorder(.data[[fill_col]], plotorder)
    )) +
    geom_col(
      width = 0.7,
      position = position_dodge(width = dodgewidth)
    ) +
    geom_text(
      aes(label = round(MEIS, 0)),
      position = position_dodge(width = dodgewidth),
      vjust = -0.5,
      size = 4.2
    ) +
    scale_fill_manual(values = c(fhiblue3, fhiblue1b)) +
    scale_y_continuous(
      breaks = seq(0, 100, 10),
      limits = c(0, ymax + 1),
      expand = expansion(mult = c(0, 0))
    ) +
    facet_wrap(~reorder(vartext, varorder)) +
    labs(
      y = ylab,
      x = "",
      fill = "",
      caption = c(label_metadata, label_kilde)
    ) +
    theme_cowplot() +
    theme_OVP
  
  # -------------------------------------------------------
  # Legg til symboler for manglende data
  # -------------------------------------------------------
  if (include_missing) {
    
    facet_missing <- d %>% filter(!is.na(prikk))
    
    p <- p +
      geom_point(
        aes(
          y = prikk,
          group = interaction(reorder(.data[[fill_col]], plotorder), kjonn_text)
        ),
        color = "black",
        size = 4.1,
        position = position_dodge(width = dodgewidth),
        show.legend = FALSE
      ) +
      geom_text(
        data = facet_missing,
        aes(label = misstext),
        x = 2.1,
        y = ymax - 3,
        size = 4,
        check_overlap = TRUE,
        inherit.aes = FALSE
      ) +
      geom_point(
        data = facet_missing,
        x = 0.83,
        y = ymax - 1.5,
        size = 3.9,
        inherit.aes = FALSE
      )
  }
  
  return(p)
}

ungdata_recentyear_plot %>% 
  filter(plotid == 1103) %>% 
  do_plot(., include_missing = FALSE)
ggsave("plots/Figur3_rus_kommmune.png", height = 5, width = 7, bg = "white")

ungdata_recentyear_plot %>%
  filter(plotid == 3435) %>%
  do_plot(., include_missing = TRUE)
ggsave("plots/Figur3_rus_kommmune_missing.png", height = 5, width = 7, bg = "white")

ungdata_recentyear_plot %>% 
  filter(plotid == 1514) %>% 
  do_plot(., missing_kommune = TRUE)
ggsave("plots/Figur3_rus_allmiss_plotfylke.png", height = 5, width = 7, bg = "white")
```



## Missing

```{r}
allmiss <- ungdata_recentyear_plot %>%
  filter(GEO_type == "kommune") %>%          
  group_by(plotid, GEO) %>%
  summarise(sumMEIS = sum(MEIS, na.rm = TRUE)) %>%
  filter(sumMEIS == 0) %>%
  pull(plotid)

length(allmiss)

missing <- select_prikk(ungdata_recentyear_plot)
missing <- setdiff(missing, allmiss)

length(missing)

length(missing)+length(allmiss)

```


# MULTIPLOT

```{r}
# Figure directory name and figure position
dirname <- "Figure3"
figposnr <- "4"


# Generate plots without missing values
plots <- ungdata_recentyear_plot %>%
  filter(!plotid %in% c(missing, allmiss)) %>%  # Exclude missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., include_missing = FALSE))  # Normal plot

# Generate plots with missing values
plots_prikk <- ungdata_recentyear_plot %>%
  filter(plotid %in% missing) %>%  # Only include missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., include_missing = TRUE))  # Include missing-value markers

# Generate plots with ALL missing values
plots_fylke <- ungdata_recentyear_plot %>%
  filter(plotid %in% allmiss) %>%  # Only include missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., missing_kommune = TRUE))  # Include missing-value markers

# Combine both lists of plots
all_plots <- c(plots, plots_prikk, plots_fylke)

# Generate file paths using `outdir` and `figposnr`
paths <- stringr::str_c(names(all_plots), "_", figposnr, "_tema", ".png")

# Create the output directory based on `outdir`
res_dir <- paste0("plots/", dirname)
dir.create(res_dir, recursive = TRUE)

# Save plots
pwalk(list(paths, all_plots), ~ggsave(filename = .x, plot = .y, 
                                      height = 5, width = 7, dpi = 200, bg = "white", 
                                      path = res_dir))

length(all_plots)
```


