---
title: "Ungdatafigur 2026"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
#options(scipen = 2)
options(digits = 2)
options(scipen=999)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggrepel)

```

O:/Prosjekt/FHP/
C:/Users/HANN/OneDrive - Helsedirektoratet/

## Functions

local functions:
  specify_decimal()
  %notint%
  select_prikk()
  map_levels()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```


## GEO  
recent year ungdata
```{r}
stednavn_GEO_BKFL_GEO2024 <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_BYDEL.csv", delim = ";")

recentyear_KH2026 <-  read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/recentyear_KH2026_KB.csv", delim = ";")
```

```{r}
recentyear_KH2026 %>% 
  group_by(AAR_end, GEO_type, completed) %>% 
  count() %>% 
  pivot_wider(names_from = GEO_type, values_from = n)
```

```{r}
recentyear_KH2026 %>% 
  filter(AAR_end == 2023)

```


## Data
ungdata indikatorer
O:/Prosjekt/FHP/
O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2025
```{r}
# select the right data
#data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2025/"  # path to the data
#files_1 <- dir(data_path, pattern = ".*UNGD.*.csv") # get file names

data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/"  # path to the data
files <- dir(data_path, pattern = ".*UNGDATA.*.csv") # get file names

# pull all selected kuber from DATERT/csv
#data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/DATERT/csv/"  # path to the data

ungdata <- data_frame(filename = files) %>% # create a data frame
                                         # holding the file names
  mutate(file_contents = map(filename,          # read files into
           ~ read_delim(file.path(data_path, .), delim = ";")) # a new data column
        )  %>% 
  unnest() %>% 
  separate(filename, into = c("filgruppe"), sep = "_202", extra = "drop", remove = TRUE) %>%
  mutate(filgruppe = gsub("_UNGD.*", "", filgruppe)) %>%
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL_GEO2024)

ungdata

#ungdata_kube <- ungdata
```

```{r}
map_levels_bycols(ungdata, bycols = "filgruppe")
```


# colors

FHI colors
```{r}
fhiblue3 <- "#393c61"
fhiblue2 <- "#0975B5"
fhiblue2b <- "#0580A7"
fhiblue1 <- "#2EA1C0"
fhiblue1b <- "#65ADC3"
fhigrey <- "#A9B7BF"
fhigreen <- "#709900"
fhidarkgreen <- "#4d7e13"
fhilightgreen <- "#aed267"
fhilightgreenman <- "#a2c24c"
fhired <- "#B11643"
fhicoral <- "#FC5F56"
fhiyellow <- "#F7B500"

show_col(c(fhiblue1, fhiblue2, fhiblue3,
           fhigrey, fhigreen, fhiblue3,
           fhired, fhicoral, fhiyellow),
         ncol = 3)

ungcolors <- c(fhigreen, fhiblue2, fhiblue1b, fhiblue3, fhiyellow, fhired)
show_col(ungcolors)
```

# Ungdata prep

```{r}
names(ungdata)
```

```{r}
map_levels(ungdata)
```

```{r}
ungdata %>% 
  group_by(filgruppe, ANTALL_GANGER, NIVA_PLAGET, FORNOYDHET, STATUS) %>% 
  count()
```


```{r}
ungdata %>% 
  group_by(filgruppe, ANTALL_GANGER, NIVA_PLAGET, FORNOYDHET, STATUS) %>% 
  count() %>% 
  select(-n) %>% 
  pivot_longer(names_to = "cat", values_to = "alt", ANTALL_GANGER:STATUS) %>% 
  filter(!is.na(alt)) %>% 
  group_by(filgruppe) %>% 
  filter(n()>1)
```

## Select indikators and filtering

```{r}
ungdata_prep <- ungdata %>% 
  filter(FORNOYDHET %in% c("LittellerSvaertFornoyd", NA_character_),
         SOES == 0,
         ) %>% 
  mutate(kjonn_text = case_when(KJONN == 0 ~ "Total",
                                KJONN == 1 ~ "Gutter",
                                KJONN == 2 ~ "Jenter"),
         soes_text = case_when(SOES == 0 ~ "Total",
                               SOES == 1 ~ "Lav",
                               SOES == 2 ~ "Middels",
                               SOES == 3 ~ "Høy"))

```


# Lokalmiljø figur

```{r}
ungdata_prep %>% 
  map_levels_bycols("filgruppe")
```

```{r}
ungdata_lokal <- ungdata_prep %>% 
  filter(filgruppe %in% c("FRITIDSORG_10aar", "SKULKETSKOLEN", "DEPRESJON", "LOKALTILBUD", "TRYGGHET", "HASJTILBUD"))

```


```{r}
ungdata_lokal %>% 
  map_levels(.)
```


```{r}
ungdata_lokal_plot_LKB <- left_join(jointable, ungdata_lokal) %>% 
  filter(SOES == 0,
         KJONN == 0) %>% 
  mutate(varorder = case_when(grepl("FRITID", filgruppe) ~ 4,
                             grepl("LOKALTILBUD", filgruppe) ~ 3,
                             #grepl("LOKALMIL", filgruppe) ~ 2,
                             grepl("DEPRESJON", filgruppe) ~ 7,
                             grepl("TRYGGHET", filgruppe) ~ 1,
                             grepl("SKULKETSKOLEN", filgruppe) ~ 5,
                             grepl("HASJ", filgruppe) ~ 6,
                             TRUE ~ NA_real_
                             ),
        vartext = case_when(grepl("FRITID", filgruppe) ~ "Organisert fritidsaktivitet",
                               grepl("LOKALTILBUD", filgruppe) ~ "Fornøyd treffsteder",
                               #grepl("LOKALMIL", filgruppe) ~ "Fornøyd lokalmiljø",
                               grepl("DEPRESJON", filgruppe) ~ "Psykiske plager",
                               grepl("TRYGGHET", filgruppe) ~ "Trygghet i nærmiljøet",
                               grepl("SKULKETSKOLEN", filgruppe) ~ "Skulket skolen",
                               grepl("HASJ", filgruppe) ~ "Hasjtilbud",
                               )) %>% 
  mutate(plotid = ifelse(grepl("^301", plotid), paste0("0", plotid), plotid))
```

missing?
```{r}
select_prikk(ungdata_lokal_plot_LKB, GEO_type = "bydel")
```
```{r}
ungdata_lokal_plot_LKB
```

```{r}
metadata
```



## Ungdata lineplot
```{r fig.height=5, fig.width=7}

# label strings
metadata <- sub(pattern = ".csv", "", basename(files)[grepl("FRITIDSORG|FORTROLIG|LOKAL|TRYGGHET|DEPRESJON|HASJTILBUD", basename(files))])

label_metadata <- paste0(metadata, collapse = "\n")

ungcolors_vartext <- c(
  "Organisert fritidsaktivitet" = fhigreen, 
  #"Fornøyd lokalmiljø" = fhiblue2, 
  "Fornøyd treffsteder" = fhiblue1b, 
  "Trygghet i nærmiljøet" = fhiblue3, 
  "Psykiske plager" = fhicoral,
  "Skulket skolen" = fhiyellow, 
  "Hasjtilbud" = fhired)

stip_text <- "Stiplet linje viser tenkt utvikling mellom gjennomførte ungdataundersøkelser i kommunen"

# min max
minmaxAAR <- ungdata_lokal_plot_LKB %>% 
  summarise(min = min(AAR_end, na.rm = TRUE),
            max = max(AAR_end, na.rm = TRUE))
minmaxMEIS <- ungdata_lokal_plot_LKB %>% 
  summarise(min = min(MEIS, na.rm = TRUE),
            max = max(MEIS, na.rm = TRUE))

# set text size
title_size <- 14
y_title_size <- 13.5      
y_text_size  <- 12      
x_text_size  <- 11      
legend_size  <- 13    
label_size   <- 3.8
ylab <- "Prosent ungdomsskoleelever, standardisert"
kildelab <- "Kilde: Ungdata"
misstext_fylke <- "Manglende eller utilstrekkelig tallgrunnlag for kommunen, fylkestall vises."


# theme
theme_ovp_lineplot <-   theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = legend_size,
                                   margin = margin(l = 2, unit = "pt"),
                                   ),
        legend.justification = "left",
        #legend.key.spacing.x = unit(-0.5, "cm"),
        plot.title = element_text(hjust = 0.45, size = title_size),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = y_title_size),
        axis.text.x = element_text(size = x_text_size),
        axis.text.y = element_text(size = y_text_size),
        plot.caption = element_text(hjust=c(1), size = c(10)),
        panel.background = element_rect(color = "white"), 
        panel.border = element_rect(linetype = "solid", fill = NA, color = "black"),
        plot.background = element_rect(fill = "white"),
        panel.grid.major.y = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.y = element_line(color = "grey", size = 0.1, linetype = 2),
        )


lineplot <- function(data, include_missing = FALSE) {

  recentyearMEIS <- data %>% 
  filter(GEO_type == "bydel",
         !is.na(MEIS)) %>% 
  group_by(Sted, varorder) %>% 
  filter(AAR_end == max(AAR_end))
  
  p <- ggplot(data = data %>% 
                filter(GEO_type == "bydel",
                       !is.na(MEIS)), 
         aes(x = AAR_end, 
             y = MEIS,
             color = reorder(vartext, varorder),
             group = interaction(Sted, filgruppe),
             )) +
  annotate("rect", xmin = 2020.4, xmax = 2021.6, ymin = 0, ymax = 100, alpha = .1) +
  annotate("text", x = 2021, y = 91, label = "Data innhentet\nunder COVID-19\npandemien", size = 2.7, color = "grey50") +
  #annotate("text", label = paste0("Tall for ", data$Sted), x = 2017, y = 110, check_overlap = TRUE) +
  geom_text(aes(label = label_metadata,
                 x = 2011.7, y = Inf),
                 hjust = 0,
                vjust = 1.2,
                 size = 1.5, 
                 color = "grey90") +
  geom_line(size = 1.1,
            alpha = 0.4,
            linetype = "dotted",
            show.legend = FALSE)  +
  geom_point(size = 4.3, alpha = 0.8) +
  geom_label_repel(data = recentyearMEIS, aes(x = AAR_end, 
             y = MEIS, label = paste0(round(MEIS,0), "%"),
             #color = reorder(vartext, varorder),
             group = interaction(Sted, filgruppe),
             ),
            nudge_x = 1.1,
            color = "black",
            size = label_size,
            direction = "y") +
  annotate("text", label = stip_text, 
           x = 2013, y = 3, size = 3, hjust = 0, color = "grey50") +
    expand_limits(y = 100) +
  scale_y_continuous(breaks = seq(0,100,20),
                     limits = c(0,100), expand = expansion(mult = c(0, 0))) +
  scale_x_continuous(breaks = seq(2012, minmaxAAR$max, 1),
                     limits = c(2011.6,minmaxAAR$max+1.1), expand = c(0,0),
                     #expansion(mult = c(0, 0)),
                     ) + 
  coord_cartesian(clip = "off") +
  scale_color_manual(values = ungcolors_vartext) +
  theme_cowplot() +
  theme_ovp_lineplot +
    guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  labs(title = paste0("Tall for ", data$Sted),
       caption = kildelab,
       y = ylab)
  
     # If missing values should be visualized, add these layers
  if (include_missing) {
    p <- p +
    annotate("text", 
           label = misstext_fylke,
           x = 2012, y = 4,
           hjust = 0, size = 3.3)

  }
  
  return(p)
}


ungdata_lokal_plot_LKB %>% 
  filter(plotid == "030105") %>% 
   lineplot(.)
ggsave("plots/Figur5_bydel.png", height = 5, width = 7, bg = "white")

ungdata_lokal_plot_LKB %>% 
  filter(plotid == 110308) %>% 
   lineplot(., include_missing = TRUE)
#ggsave("plots/fritidpsyk_kommune.png", height = 5, width = 7, bg = "white")
```


NO all miss

Show fylke/landtall for missing last 5 years
```{r}
allmiss <- ungdata_lokal_plot_LKB %>% 
  filter(GEO_type == "bydel",
         AAR_end >= minmaxAAR$max-5) %>% 
  group_by(plotid) %>% 
  summarise(sumRATE = sum(MEIS, na.rm = TRUE)) %>% 
  filter(sumRATE == 0) %>% 
  pull(plotid)

length(allmiss)

allmiss
```


```{r}
# Figure directory name and figure position
dirname <- "Figure5_BYDEL"
figposnr <- "6"


# Generate plots without missing values
plots <- ungdata_lokal_plot_LKB %>%
  filter(!plotid %in% allmiss) %>%  # Exclude missing-value plots
  split(.$plotid) %>%
  map(~lineplot(., include_missing = FALSE))  # Normal plot

# Generate plots with missing values
plots_prikk <- ungdata_lokal_plot_LKB %>%
  filter(plotid %in% allmiss) %>%  # Only include missing-value plots
  split(.$plotid) %>%
  map(~lineplot(., include_missing = TRUE))  # Include missing-value markers

# Combine both lists of plots
all_plots <- c(plots, plots_prikk)

# Generate file paths using `outdir` and `figposnr`
paths <- stringr::str_c(names(all_plots), "_", figposnr, "_tema", ".png")

# Create the output directory based on `outdir`
res_dir <- paste0("plots/", dirname)
dir.create(res_dir, recursive = TRUE)

# Save plots
pwalk(list(paths, all_plots), ~ggsave(filename = .x, plot = .y, 
                                      height = 5, width = 7, dpi = 200, bg = "white", 
                                      path = res_dir))

length(all_plots)
```


