---
title: "Ungdata - ungdomsskole + videregående rus + nikotin"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
#options(scipen = 2)
options(digits = 2)
options(scipen=999)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggrepel)

```

O:/Prosjekt/FHP/
C:/Users/HANN/OneDrive - Helsedirektoratet/

## Functions

local functions:
  specify_decimal()
  %notint%
  select_prikk()
  map_levels()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```

```{r}
fhiblue3 <- "#393c61"
fhiblue2 <- "#0975B5"
fhiblue2b <- "#0580A7"
fhiblue1 <- "#2EA1C0"
fhiblue1b <- "#65ADC3"
fhigrey <- "#A9B7BF"
fhigreen <- "#709900"
fhidarkgreen <- "#4d7e13"
fhilightgreen <- "#aed267"
fhilightgreenman <- "#a2c24c"
fhired <- "#B11643"
fhicoral <- "#FC5F56"
fhiyellow <- "#F7B500"

show_col(c(fhiblue1, fhiblue2, fhiblue3,
           fhigrey, fhigreen, fhiblue3,
           fhired, fhicoral, fhiyellow),
         ncol = 3)

```


## GEO  
recent year ungdata
```{r}
stednavn_GEO_BKFL_GEO2024 <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", delim = ";")

recentyear_KH2026 <-  read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/recentyear_KH2026_KB.csv", delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_GEO2024.csv", delim = ";")

```


```{r}
jointable
```

```{r}
recentyear_KH2026
```


# Data

## Data ungdomsskole

```{r}

data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/"  # path to the data
files <- dir(data_path, pattern = ".*UNGDATA_20.*\\.csv") # get file names

uskole <- data_frame(filename = files) %>% # create a data frame
                                         # holding the file names
  mutate(file_contents = map(filename,          # read files into
           ~ read_delim(file.path(data_path, .), delim = ";")) # a new data column
        )  %>% 
  unnest() %>% 
  separate(filename, into = c("filgruppe"), sep = "_202", extra = "drop", remove = TRUE) %>%
  mutate(filgruppe = gsub("_UNGD.*", "", filgruppe)) %>%
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL_GEO2024)

metadata_uskole <- sub(pattern = ".csv", "", basename(files)[grepl("ROYK|SNUS|VAPE|HASJBRUK|ANDRENARKO|ALKO", basename(files))])

```


```{r}
uskole_prep <- uskole %>% 
  filter(grepl("ALKO|HASJBRUK|NARKO|VAPE|ROYK|SNUS", filgruppe),
         SOES == 0,
         ) %>% 
  select(-c(DEPRESJON, FORNOYDHET, NIVA_PLAGET, HARVENN, DELTARNAA, LOKALTILBUD, LIVSKVALITET, REGELBRUDD, SKJERMTID, STATUS))
```

```{r}
map_levels_bycols(uskole_prep, bycols = "filgruppe")
```
## Data videregående

OBS!! ikke godkjente kuber (enda)

```{r}

data_path <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/"  # path to the data
files <- dir(data_path, pattern = ".*UNGDATA_VGO.*.csv")

vgo <- data_frame(filename = files) %>% # create a data frame
                                         # holding the file names
  mutate(file_contents = map(filename,          # read files into
           ~ read_delim(file.path(data_path, .), delim = ";")) # a new data column
        )  %>% 
  unnest() %>% 
  separate(filename, into = c("filgruppe"), sep = "_202", extra = "drop", remove = TRUE) %>%
  mutate(filgruppe = gsub("_UNGD.*", "", filgruppe)) %>%
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL_GEO2024)

metadata_vgo <- sub(pattern = ".csv", "", basename(files)[grepl("ROYK|SNUS|VAPE|HASJBRUK|ANDRENARKO", basename(files))])

```

```{r}
map_levels_bycols(vgo, bycols = "filgruppe")
```
## Bind
```{r}
names(uskole_prep)
names(vgo)

uskolevgo <- rbind(uskole_prep %>% 
                     mutate(ALDER = "Ungdomsskole"), 
                   vgo %>% 
                     mutate(ALDER = "Videregående"))

```

```{r}

metadata <- c(metadata_uskole, metadata_vgo)

```


# Plot

REMOVE UNGDOMSSKOLE ANDRENARKO < 2021 !!!!

```{r}
uskolevgo_plot <- uskolevgo %>% 
  filter(KJONN %in% c(1,2),
         SOES == 0,
         GEO == 0, 
         AAR_end >= 2015) %>% 
    mutate(MEIS = ifelse(filgruppe == "ANDRENARKO" & ALDER == "Ungdomsskole" & AAR_end < 2021, NA_real_, MEIS),
           kjonn_text = case_when(KJONN == 0 ~ "Total",
                                KJONN == 1 ~ "Gutter",
                                KJONN == 2 ~ "Jenter"),
           vartext = case_when(grepl("ROYK", filgruppe) ~ "Bruker røyk \nukentlig/daglig",
                               grepl("SNUS", filgruppe) ~ "Bruker snus \nukentlig/daglig",
                               grepl("VAPE", filgruppe) ~ "Bruker vape \nukentlig/daglig",
                               grepl("ALKOHOL", filgruppe) ~ "Har vært \nberuset på alkohol",
                               grepl("HASJBRUK", filgruppe) ~ "Har brukt \nhasj/cannabis",
                               grepl("ANDRENARKO", filgruppe) ~ "Har brukt andre \nnarkotiske stoff",
                               ),
           alder_aar = case_when(ALDER == "Ungdomsskole" ~ paste0(ALDER, " (13-16 år)"),
                                 ALDER == "Videregående" ~ paste0(ALDER, " (16-19 år)")),
           group = interaction(alder_aar, kjonn_text, sep = ", "),
           )
```

```{r}
uskolevgo_plot %>% 
  map_levels()

uskolevgo_plot %>% 
  count(group, alder_aar, ALDER, kjonn_text)
```


(alle) Rusmidler + nikotin
```{r fig.height=5, fig.width=7}

# set text size
title_size <- 14
y_title_size <- 14      
y_text_size  <- 11      
x_text_size  <- 10
facet_title_size <- 13.5
legend_size  <- 14  
label_size   <- 3.5
ymax <- 25
ylab <- "Prosent ungdom, standardisert"

theme_facetline <- theme(axis.title.x = element_blank(),
                         axis.title.y = element_text(size = y_title_size),
                         legend.text = element_text(size = legend_size),
                          legend.position = "bottom",
                          legend.title = element_blank(),
                         legend.box.spacing = unit(0.2, "cm"),
                         legend.key.spacing = unit(0.1, "cm"),
                         strip.text = element_text(size = facet_title_size),
                         strip.background = element_rect(fill = "white"), 
                          panel.background = element_rect(color = "white"), 
                          panel.border = element_rect(linetype = "solid", fill = NA, color = "black"),
                         panel.spacing = unit(0.1, "lines"),
                          plot.background = element_rect(fill = "white"),
                          plot.margin = margin(t = 1, r = 10, b = 0, l = 0, unit = "pt"),
                          panel.grid.major.x = element_line(color = "grey", size = 0.1, linetype = 1),
                          panel.grid.minor.x = element_line(color = "grey", size = 0.1, linetype = 2),
                          panel.grid.major.y = element_line(color = "grey", size = 0.1, linetype = 1),
                          panel.grid.minor.y = element_line(color = "grey", size = 0.1, linetype = 2),
        )

uskolevgo_plot %>%
  filter(!is.na(MEIS)) %>% 
  ggplot(aes(x = AAR_end,
             y = MEIS,
             color = group,
             group = interaction(group, filgruppe)),
             ) +
  geom_line(size = 1.3) +
  #geom_point(size = 1.5) +
  annotate("rect", xmin = 2020.4, xmax = 2021.6, ymin = 0, ymax = ymax, alpha = 0.1, fill = "grey20") +
  annotate("text", x = 2021, y = 23.5, label = "COV19", size = 2, color = "grey30") +
#scale_x_continuous(breaks = seq(minmaxAAR$min, minmaxAAR$max, 2)) +
  scale_color_manual(values = c(fhiblue2, fhiblue3, fhicoral, fhired)) +
  scale_x_continuous(breaks = seq(2017, 2025, 4)) +
  scale_y_continuous(limits = c(0,ymax), expand = c(0,0)) +
  #facet_wrap(~vartext, scales = "free") +
  facet_wrap(~vartext) +
  labs(y = ylab) +
  theme_cowplot() +
  theme_facetline +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))
ggsave("plots/Figur1_uskole_vgo_rusnikotin.png", height = 5, width = 7)
```
O:/Prosjekt/FHP/PRODUKSJON/VALIDERING/Temafigurer
```{r}
metadata
write_delim(as.data.frame(metadata), "plots/OVP2026_Figur1_Kuber_metadata.txt", delim = "\n")
write_delim(as.data.frame(metadata), "O:/Prosjekt/FHP/PRODUKSJON/VALIDERING/Temafigurer/OVP2026_Figur1_Kuber_metadata.txt", delim = "\n")
```



####################

```{r fig.height=5, fig.width=7}
uskolevgo_plot %>%
  filter(!is.na(MEIS)) %>% 
  ggplot(aes(x = AAR_end,
             y = MEIS,
             color = group,
             group = interaction(group, filgruppe)),
             ) +
  geom_line() +
  geom_point() +
  #scale_x_continuous(breaks = seq(minmaxAAR$min, minmaxAAR$max, 2)) +
  scale_color_manual(values = c(fhiblue2, fhiblue3, fhicoral, fhired)) +
  scale_x_continuous(breaks = seq(2013, 2025, 4)) +
  facet_wrap(~vartext, scales = "free_y") +
  labs(y = "Andel ungdom i prosent") +
  theme_cowplot() +
  theme(axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        panel.background = element_rect(color = "white"), 
        panel.border = element_rect(linetype = "solid", fill = NA, color = "black"),
        plot.background = element_rect(fill = "white"),
        panel.grid.major.x = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.x = element_line(color = "grey", size = 0.1, linetype = 2),
        panel.grid.major.y = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.y = element_line(color = "grey", size = 0.1, linetype = 2),
        ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))
#ggsave("plots/uskole_vgo_rusnik_tid.png", height = 5, width = 7)
```


Trying to set two different y limits
```{r}

# Build a named vector for labeller: names = filgruppe codes, values = vartext
label_map <- uskolevgo_plot %>%
  distinct(filgruppe, vartext) %>%
  { setNames(.$vartext, .$filgruppe) }

uskolevgo_plot %>%
  ggplot(aes(x = AAR_end,
             y = MEIS,
             color = group,
             group = interaction(group, filgruppe))) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(2013, 2025, 4)) +
  scale_color_manual(values = c(fhiblue2, fhiblue3, fhicoral, fhired)) +
  facet_wrap(~ filgruppe, labeller = labeller(filgruppe = label_map)) +
  facetted_pos_scales(
    y = list(
      vars(filgruppe == "ALKOHOL") ~ scale_y_continuous(limits = c(0, 65)),
      vars(filgruppe != "ALKOHOL") ~ scale_y_continuous(limits = c(0, 22))
    )
  ) +
  labs(y = "Andel ungdom i prosent") +
  theme_cowplot() +
  theme(axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        panel.background = element_rect(color = "white"), 
        panel.border = element_rect(linetype = "solid", fill = NA, color = "black"),
        plot.background = element_rect(fill = "white"),
        panel.grid.major.x = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.x = element_line(color = "grey", size = 0.1, linetype = 2),
        panel.grid.major.y = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.y = element_line(color = "grey", size = 0.1, linetype = 2),
        ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))
```


```{r fig.height=5, fig.width=7}
uskolevgo_plot %>% 
  filter(grepl("HASJ|NARKO|SNUS|VAPE", filgruppe)) %>% 
  ggplot(aes(x = AAR_end,
             y = MEIS,
             color = group,
             group = interaction(group, filgruppe)),
             ) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(minmaxAAR$min, minmaxAAR$max, 2)) +
  scale_color_manual(values = c(fhiblue2, fhiblue3, fhicoral, fhired)) +
  scale_x_continuous(breaks = seq(2013, 2025, 2)) +
  scale_y_continuous(limits = c(0,24)) +
  facet_wrap(~vartext, scales = "free_y") +
  labs(y = "Andel ungdom i prosent") +
  theme_cowplot() +
  theme(axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        panel.background = element_rect(color = "white"), 
        panel.border = element_rect(linetype = "solid", fill = NA, color = "black"),
        plot.background = element_rect(fill = "white"),
        panel.grid.major.x = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.x = element_line(color = "grey", size = 0.1, linetype = 2),
        panel.grid.major.y = element_line(color = "grey", size = 0.1, linetype = 1),
        panel.grid.minor.y = element_line(color = "grey", size = 0.1, linetype = 2),
        ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))
ggsave("plots/uskole_vgo_narkonicotin_tid.png", height = 5, width = 7)
```

```{r}

```




```{r}
uskolevgo_plot %>% 
  group_by(filgruppe, ALDER) %>% 
  minmax(.)
```


```{r}
vgo %>% 
  arrange(-MEIS)
```


```{r}
vgo %>% 
  filter(GEO == 3,
         SOES == 0) %>% 
  select(filgruppe, GEO, AAR, KJONN, SOES, MEIS) %>% 
  arrange(filgruppe, KJONN, AAR)
```

```{r}
vgo %>% 
  filter(GEO == 0,
         SOES == 0) %>% 
  select(filgruppe, GEO, AAR, KJONN, SOES, MEIS) %>% 
  arrange(filgruppe, KJONN, AAR)
```

```{r fig.height=6, fig.width=10}
vgo %>% 
  filter(GEO == 0,
         SOES == 0) %>% 
  ggplot(aes(x = AAR_end,
             y = MEIS,
             color = as.factor(KJONN))) +
  geom_line(size = 2) +
  scale_x_continuous(breaks = c(seq(2016, 2025, 2))) +
  theme(legend.position = "bottom") +
  facet_wrap(~filgruppe)
```

```{r}
map_levels(uskolevgo_plot)
```





