---
title: "R Notebook"
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
options(scipen = 999)

#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggbeeswarm, ggimage)

```


Figur 3 (C8verst til hC8yre side tre):
barn (0-18 C%r), eldre (80 +) og andel i lavinntektshus. (alle aldre). 


## Functions

local functions:
  specify_decimal()
  %notin%
  select_prikk()
  map_levels()
  minmax()
  minmax_bycols()
  minmax_bycolname()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```

## Data

GEO
Jointable
recent year

C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data
```{r}
stednavn_GEO_BKFL <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", 
                                delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_fylke_2024.csv", delim = ";")

```


KUBER

```{r}
#file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/KOMMUNEHELSA/KH2025NESSTAR/syssel_GK_2024-06-17-14-13.csv"
file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/SYSSELSATTE_UTDANN_1_2026-02-06-09-17.csv"
syssel <- data.table::fread(file = file)
syssel_data <- gsub(".csv", "", basename(file))

file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/NEET_UTDANN_2026-02-09-13-34.csv"
neet <- data.table::fread(file = file)
neet_data <- gsub(".csv", "", basename(file))

file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/SYKEFRAV_2026-02-22-17-26.csv"
syke <- data.table::fread(file = file)
syke_data <- gsub(".csv", "", basename(file))
```

```{r}
data_data <- paste(neet_data,
                   syssel_data,
                   syke_data,
                   sep = "\n")
```

# Data prep

```{r}
names(syssel)
names(neet)
names(syke)
```

```{r}
map_levels(syssel)
map_levels(neet)
map_levels(syke)

```


```{r}
data <- rbind(syssel %>% 
                filter(KJONN == 0,
                       UTDANN == 0,
                       STILLINGSANDEL == "HELDELTID_TOT",
                       ALDER == "20_66") %>% 
                 select(GEO, AAR, ALDER, MEIS, TELLER, SPVFLAGG) %>% 
                 mutate(var = "syssel"),
               neet %>% 
                filter(KJONN == 0,
                       UTDANN == 0,
                       ALDER == "20_66") %>% 
                 select(GEO, AAR, ALDER, MEIS, TELLER, SPVFLAGG) %>% 
                 mutate(var = "neet"),
              syke %>% 
                filter(KJONN == 0,
                       ALDER == "20_66") %>% 
                select(GEO, AAR, ALDER, MEIS, TELLER, SPVFLAGG) %>% 
                 mutate(var = "syke")
              ) %>% 
    separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
    mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
    left_join(., stednavn_GEO_BKFL)

```

```{r}
data %>% 
  group_by(var) %>% 
  filter(AAR_end == max(AAR_end)) %>% 
  count(var, AAR)

data_filt <- data %>% 
  group_by(var) %>% 
  filter(AAR_end == max(AAR_end)) %>% 
  ungroup() %>% 
  mutate(varorder = case_when(var == "syssel" ~ 1,
                          var == "neet" ~ 2,
                          var == "syke" ~ 3,
                          TRUE ~ 0)) %>% 
  filter(varorder >= 1)

map_levels(data_filt)

```

x = c(0.12, 0.55, 0.88)

```{r}
data_plot <- data_filt %>% 
  right_join(., jointable) %>% 
  relocate(plotid, .before = GEO) %>% 
  filter(GEO_type %in% c("fylke")) %>%
  arrange(plotid) %>% 
  mutate(y_value = case_when(var == "syssel" ~ 0.78,
                             var == "neet" ~ 0.36,
                             var == "syke" ~ 0.79,
                             )) %>% 
  mutate(x_value = case_when(var == "syssel" ~ 0.325,
                             var == "neet" ~ 0.54,
                             var == "syke" ~ 0.74,
                             )) %>% 
  mutate(var_plottext = case_when(var == "syke" ~ "Sykefravær",
                                   var == "neet" ~ "NEET",
                                   var == "syssel" ~ "sysselsetting")) %>% 
  mutate(label_text = case_when(
    var %in% c("syssel", "neet") ~ paste0(number(round(TELLER,0), big.mark = " "), "\n", round(MEIS, 0), "%"),
    var == "syke" ~ paste0(round(MEIS, 1), "%"),
    )) %>% 
  mutate(prikk = ifelse(is.na(MEIS), 3, NA_real_)) %>% 
  mutate(plotid = ifelse(plotid == 3, "03", plotid)) # ensure 4 digit Oslo geo code


```

```{r}
data_plot
```

No missing
```{r}
missing <- select_prikk(data_plot)

```


# Multiplot

```{r fig.height=5, fig.width=6.5}
bg <- data.frame(x = 0.5, y = 0.5, img = "Figur_1_v06_fraværsdager.png")
titlesize <- 17
textsize  <- 6.5

do_plot <- function(data,
                    include_missing = FALSE,
                    img_size = 1.4,       # size for geom_image
                    title_size = titlesize,
                    text_size  = textsize,
                    xlim = c(0.15, 0.85),
                    ylim = c(0, 1),
                    clip = "off") {
  
  plottitle <- unique(paste0("Tall for ", data$Sted[1]))

  p <- ggplot() +
    geom_image(data = bg, aes(x, y, image = img), size = img_size, asp = 1) +
    geom_text(
      data = data,
      aes(x = x_value, y = y_value, label = label_text),
      color = "black", size = text_size, fontface = "bold"
    ) +
    coord_cartesian(xlim = xlim, ylim = ylim, expand = c(0,0), clip = clip) +
    theme_void() +
    theme(
      plot.title  = element_text(size = title_size, hjust = 0.54),
      plot.margin = margin(0, 0, 0, 0),
      plot.caption = element_text(hjust=c(0,1), size = c(5,10), color = c("grey50", "black"))
    ) +
    labs(title = plottitle,
         caption = c(data_data, "Kilde: SSB"))
        
  return(p)
}

data_plot %>% 
  filter(plotid == "11") %>% 
  do_plot(include_missing = FALSE)
ggsave("plots/Figur1_arbeidsliv_fylke.png", height = 5, width = 6.5, bg = "white", dpi = 150)

```

optimizing


missing
```{r}
SPVflagg <- select_prikk(data_plot)
SPVflagg
```


```{r}
# Figure directory name and figure position
dirname <- "Figure1_FYLKE"
figposnr <- "2"

SPVflagg <- select_prikk(data_plot)  # Get plot IDs with missing values


# Generate plots without missing values
plots <- data_plot %>%
  filter(!plotid %in% SPVflagg) %>%  # Exclude missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., include_missing = FALSE))  # Normal plot

# Generate plots with missing values
plots_prikk <- data_plot %>%
  filter(plotid %in% SPVflagg) %>%  # Only include missing-value plots
  split(.$plotid) %>%
  map(~do_plot(., include_missing = TRUE))  # Include missing-value markers

# Combine both lists of plots
all_plots <- c(plots, plots_prikk)

# Generate file paths using `outdir` and `figposnr`
paths <- stringr::str_c(names(all_plots), "_", figposnr, "_tema", ".png")

# Create the output directory based on `outdir`
res_dir <- paste0("plots/", dirname)
dir.create(res_dir, recursive = TRUE)

# Save plots
pwalk(list(paths, all_plots), ~ggsave(filename = .x, plot = .y, 
                                      height = 5, width = 6.5, dpi = 200, bg = "white", 
                                      path = res_dir))

length(all_plots)
```


# Landstall til nettartikkel

```{r}
data_plot_landet <- data_filt %>% 
  filter(GEO_type %in% c("land")) %>%
  mutate(plotid = GEO) %>% 
  mutate(y_value = case_when(var == "syssel" ~ 0.78,
                             var == "neet" ~ 0.36,
                             var == "syke" ~ 0.79,
                             )) %>% 
  mutate(x_value = case_when(var == "syssel" ~ 0.325,
                             var == "neet" ~ 0.54,
                             var == "syke" ~ 0.74,
                             )) %>% 
  mutate(var_plottext = case_when(var == "syke" ~ "Sykefravær",
                                   var == "neet" ~ "NEET",
                                   var == "syssel" ~ "sysselsetting")) %>% 
  mutate(label_text = case_when(
    var %in% c("syssel", "neet") ~ paste0(number(round(TELLER,0), big.mark = " "), "\n", round(MEIS, 0), "%"),
    var == "syke" ~ paste0(round(MEIS, 0), "%"),
    )) %>% 
  mutate(prikk = ifelse(is.na(MEIS), 3, NA_real_)) %>% 
  mutate(plotid = ifelse(plotid == 301, "0301", plotid)) # ensure 4 digit Oslo geo code

data_plot_landet
```

## plot
```{r fig.height=5, fig.width=6.5}
bg <- data.frame(x = 0.5, y = 0.5, img = "Figur_1_v06_fraværsdager.png")
titlesize <- 17
textsize  <- 6.5

do_plot_landet <- function(data,
                    include_missing = FALSE,
                    img_size = 1.4,       # size for geom_image
                    title_size = titlesize,
                    text_size  = textsize,
                    xlim = c(0.15, 0.85),
                    ylim = c(0, 1),
                    clip = "off") {
  
  plottitle <- unique(paste0("Tall for ", data$Sted[1]))

  p <- ggplot() +
    geom_image(data = bg, aes(x, y, image = img), size = img_size, asp = 1) +
    geom_text(
      data = data,
      aes(x = x_value, y = y_value, label = label_text),
      color = "black", size = text_size, fontface = "bold"
    ) +
    coord_cartesian(xlim = xlim, ylim = ylim, expand = c(0,0), clip = clip) +
    theme_void() +
    theme(
      plot.title  = element_text(size = title_size, hjust = 0.54),
      plot.margin = margin(0, 0, 0, 0),
      plot.caption = element_text(hjust=c(0,1), size = c(5,10), color = c("grey50", "black"))
    ) +
    labs(title = plottitle,
         caption = c(data_data, "Kilde: SSB"))
        
  return(p)
}


data_plot_landet %>% 
  filter(plotid == 0) %>% 
  do_plot_landet(.)
ggsave("../Nettartikkel/Figur1_arbeidsliv.png", height = 5, width = 6.5, bg = "white", dpi = 150)
```



