---
title: "Figur 5 - NEET"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggbeeswarm)

```


## Functions

local functions:
  specify_decimal()
  %notin%
  select_prikk()
  map_levels()
  minmax()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```

## Data

GEO
Jointable
recent year

C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data
```{r}
stednavn_GEO_BKFL <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_BYDEL.csv", delim = ";")



```


KUBER

O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER
```{r}
file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/NEET_UTDANN_2026-02-09-13-34.csv"
neet <- read_delim(file, delim = ";") %>% 
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL)
neet_data <- gsub(".csv", "", basename(file))

```

Hdir colors
```{r}
hdirpink3 <- "#7C145C"
hdirpink2 <- "#DBB5ED"
hdirpink1 <- "#FCF3F9"
hdiryellow3 <- "#FFBF00"
hdiryellow2 <- "#FFE187"
hdiryellow1 <- "#FFF9E6"

hdirblue2b <- "#048eb7"
hdirblue_f4 <- "#047FA4"

hdirfarge1 <- "#025169"
hdirfarge2 <- "#0069E8"
hdirfarge3 <- "#7C145C"
hdirfarge4 <- "#047FA4"
hdirfarge5 <- "#C68803"
hdirfarge6 <- "#38A389"
hdirfarge7 <- "#6996CE"
hdirfarge8 <- "#366558"
hdirfarge9 <- "#BF78DE"
hdirfarge10 <- "#767676"

colors_utdann <- c(hdirfarge3, hdirfarge5, hdirfarge9)
colors_utdann2 <- c(hdirfarge4, hdirfarge5, hdirfarge9)
colors_utdann3 <- c(hdirfarge3, hdirfarge5, hdirfarge4)
show_col(c(colors_utdann,
           colors_utdann2,
           colors_utdann3),
         ncol = 3)

fhigrey <- "#A9B7BF"

```

https://designsystem.helsedir.no/0410c9e94/p/736ae4-farger
```{r}
hdirblue3 <- "#025169"
hdirblue2 <- "#047FA4"
hdirblue1 <- "#8CE2FD"
hdirblue1b <- "#9ce0f5"
hdirblue1c <- "#83cfe6"
fhiblue1b <- "#65ADC3"
hdirgreen3 <- "#366558"
hdirgreen2 <- "#95DBCA"
hdirgreen2b <- "#78d1bb"
hdirgreen_f6 <- "#38A389"
hdirgreen1b <- "#c5e9df"
hdirgreen1c <- "#dff7f0"
hdirgreen1d <- "#cff3e9"
hdirgreen1 <- "#EBF5F2"
hdirwhite <- "#FFFFFF"

show_col(c(hdirblue3, hdirblue2, hdirblue1,
           hdirblue3, hdirblue2, hdirblue1b,
           hdirblue3, hdirblue2, hdirblue1c
           ),
         ncol = 3)

show_col(c(hdirblue3, hdirgreen2, hdirgreen1b,
           hdirblue3, hdirgreen2, hdirgreen1d,
           hdirblue3, hdirgreen2, hdirgreen1c,
           hdirblue3, hdirgreen2, hdirgreen1,
           hdirblue3, hdirgreen2b, hdirgreen1d
           ),
         ncol = 3)

show_col(c(hdirblue3, hdirblue2, hdirgreen1d,
           hdirblue3, hdirblue2, hdirgreen1b,
           hdirblue3, hdirblue2, hdirgreen2
           ),
         ncol = 3)

colors_LFK <- c(hdirblue3, hdirblue2, hdirgreen2)

colors_LK <- c(hdirblue3, hdirgreen_f6)
show_col(colors_LK)

colors_LKB <- c(hdirblue3, hdirgreen_f6, hdirgreen2)
show_col(colors_LKB, ncol = 3)
```

# Data prep


```{r}

map_levels(neet)

```

## Filtering

15-29, 30-44, 45-66 

```{r}

neet_prep <- neet %>% 
  filter(KJONN == 0,
         ALDER %in% c("15_29", "30_44", "45_66"),
         AAR_end == max(AAR_end),
         UTDANN == 0) %>% 
  mutate(ALDER_text = case_when(ALDER == "15_29" ~ "15-29 år",
                                ALDER == "30_44" ~ "30-44 år",
                                ALDER == "45_66" ~ "45-66 år",
                                ),
         ALDER_order = case_when(ALDER == "15_29" ~ 1,
                                ALDER == "30_44" ~ 2,
                                ALDER == "45_66" ~ 3,
                                ),
        )

```


```{r}
neet_plot <- left_join(jointable, neet_prep) %>% 
  filter(GEO_type %in% c("land", "kommune", "bydel")) %>% 
  mutate(prikk = ifelse(is.na(RATE), 3, NA_real_)) %>% 
  mutate(label_metadata = neet_data,
         label_kilde = "Kilde: SSB",
         ) %>% 
  mutate(plotid = ifelse(grepl("^301", plotid), paste0("0", plotid), plotid)) # ensure 6 digit Oslo geo code

map_levels(neet_plot)

```

```{r}
min_max <- minmax(neet_plot)
min_max
```


# PLOT

## Søyle facet LK


```{r fig.height=5, fig.width=7}

x_text_size <- 12
y_text_size <- 11.5
axis_title_size <- 14
legend_size <- 13.5
stripsize <- 14
colwidth <- 0.7
ymax <- 100
ylab <- "Andel standardisert (prosent)"
kildelab <- paste0("Kilde: SSB (", min_max$max_aar, ")")
#colors_LKB <- c(hdirblue3, hdirgreen_f6, hdirgreen2)
colors_LKB <- c(hdirblue3, fhigrey, hdirblue2)

theme_OVP <- theme_cowplot() +
    theme(#legend.position = "bottom",
          legend.position = "none",
          legend.text = element_text(size = legend_size),
          legend.margin=margin(t=-6),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = x_text_size,
                                     angle = 45, hjust = 1,
                                     ),
          axis.text.y = element_text(size = y_text_size),
          axis.title.y = element_text(size = axis_title_size),
          strip.background = element_blank(),
          strip.text = element_text(size = stripsize),
          line = element_blank(),
          axis.ticks.y = element_line(color = "grey50"),
          panel.border = element_rect(fill = NA, color = "grey50"),
          plot.caption = element_text(hjust=c(0,1), size = c(5,9)),
          plot.margin = margin(t = 2, r = 5, b = 2, l = 5),
          panel.background = element_rect(color = "white"), 
          )

misstext <- "Manglende eller anonymiserte tall"

do_plot <- function(data, include_missing = FALSE) {  # default is missing FALSE, no need to speficy
    
   highest_plotorder <- data %>% filter(plotorder == max(plotorder)) %>% pull(Sted)

    p <- data %>% 
    ggplot(aes(y = MEIS, x = reorder(ALDER_text, ALDER_order),
              fill = reorder(Sted, plotorder))) +
    geom_col(width = colwidth,
             position = position_dodge(width = colwidth)) +
  geom_text(aes(label = round(MEIS,0)),
            nudge_y = 4,
            size = 4) +
    scale_fill_manual(values = c(colors_LKB)) +
    #scale_x_discrete(expand = expansion(mult = c(0.20, 0.20))) +
    scale_y_continuous(breaks = seq(0,100,10),
                       limits = c(0,100), expand = expansion(mult = c(0, 0))) +
    coord_cartesian(ylim = c(0,100),
                    #xlim = c(1,3),
                    clip = "off") +
  theme_OVP +
    labs(y = ylab,
         caption = c(neet_data, kildelab),
         fill = "") +
      # facet_wrap(~reorder(Sted, plotorder)) 
    facet_grid(~reorder(Sted, plotorder), scales = "free", space = "free")
  
   
   # If missing values should be visualized, add these layers
  if (include_missing) {
    p <- p +
      geom_point(aes(y = prikk,
                     group = reorder(Sted, plotorder)), 
                     #group = reorder(Sted, plotorder)), 
                 color = "black", size = 4.1,
                 position = position_dodge(width = colwidth),
                 show.legend = FALSE) +
       geom_text(data = data %>% filter(Sted == highest_plotorder),
              aes(label = misstext,
                     group = reorder(Sted, plotorder)),
                 x = 1,
                 y = 97,
                 hjust = 0,
                 size = 3.9,
                 check_overlap = TRUE,
                 inherit.aes = FALSE) +
       geom_point(data = data %>% filter(Sted == highest_plotorder),
              aes(group = reorder(Sted, plotorder)),
                  x = 0.8, 
                  y = 96.6,
                  size = 4,
                  inherit.aes = FALSE)
  }
  
  return(p)
}

neet_plot %>% 
  filter(#GEO_type %in% c("land", "kommune"),
         ) %>% 
  filter(plotid == "030101") %>% 
  do_plot(., include_missing = FALSE)
ggsave("plots/neet_UTDANN_KJONN_bydel.png", height = 5, width = 7, bg = "White")

# neet_plot %>%
#   filter(#GEO_type %in% c("land", "kommune"),
#          ) %>%
#   filter(plotid == 1151) %>%
#   do_plot(., include_missing = TRUE)

#ggsave("plots/neet_UTDANN_KJONN_kommune.png", height = 5, width = 7, bg = "White")
```


no missing for bydel
```{r}
select_prikk(neet_plot)
```



# MULTIPLOT

length(nomiss)+length(kjonnmiss)+length(kjonnmiss3more)+length(kjonntotmiss)

```{r}
missing <- select_prikk(neet_plot)

dirname <- "Figur5_alder_BYDEL_gråblå"
figposnr <- "6"

res_dir <- paste0("plots/", dirname)
dir.create(res_dir, recursive = TRUE)

# Plot Level 1 - Plot kjønn - no missing
plots_1 <- neet_plot %>% 
  filter(plotid %notin% missing) %>%
  split(.$plotid) %>% 
  map(~do_plot(.x, include_missing = FALSE))

# Plot Level 2 - missing
plots_2 <- neet_plot %>% 
  filter(plotid %in% missing) %>%  
  split(.$plotid) %>% 
  map(~do_plot(.x, include_missing = TRUE))

# Combine both lists of plots
#all_plots <- c(plots_5)
all_plots <- c(plots_1, plots_2)

# Generate file paths using `outdir` and `figposnr`
paths <- stringr::str_c(names(all_plots), "_", figposnr, "_tema", ".png")

# Save plots
pwalk(list(paths, all_plots), ~ggsave(filename = .x, plot = .y, 
                                      height = 5, width = 7, dpi = 200, bg = "white", 
                                      path = res_dir))
length(plots_1)
length(plots_2)

length(plots_1)+length(plots_2)

length(all_plots)
```




