---
title: "Figur 4 - Sykefravær"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Hanna Helgeland
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Start with clear environment
rm(list = ls())

# install pacman - a tool for easy installing and loading the other required packages
if (!require("pacman")) install.packages("pacman")

# use pacman to install - if needed, and load packages
pacman::p_load(tidyverse, cowplot, Hmisc, ggpubr, scales, ggbeeswarm)

```


Figur 4: 
Sykefravær, legemeldt (linjediagram)
etter kjønn



## Functions

local functions:
  specify_decimal()
  %notin%
  select_prikk()
  map_levels()
  minmax()
```{r}
source("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_scripts/local_functions.R")
```

## Data

GEO
Jointable
recent year

C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data
```{r}
stednavn_GEO_BKFL <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/stednavn_GEO_BKFL_GEO2024.csv", delim = ";")

jointable <- read_delim("C:/Users/HANN/OneDrive - Helsedirektoratet/R_local/00_data/jointable_multiplot_BYDEL.csv", delim = ";")

```


KUBER

O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER
```{r}
#file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/STATBANK_2026/SYKEFRAV_2026-02-22-17-26.csv"
file <- "O:/Prosjekt/FHP/PRODUKSJON/PRODUKTER/KUBER/STATBANK/DATERT/csv/SYKEFRAV_2026-02-22-17-26.csv"
syke <- read_delim(file, delim = ";") %>% 
  separate(AAR, into = c("AAR_start", "AAR_end"), sep = "_", remove = FALSE) %>% 
  mutate_at(c("GEO", "AAR_start", "AAR_end"), as.integer) %>% 
  left_join(., stednavn_GEO_BKFL)
syke_data <- gsub(".csv", "", basename(file))

```

Hdir colors
```{r}
hdirpink3 <- "#7C145C"
hdirpink2 <- "#DBB5ED"
hdirpink1 <- "#FCF3F9"
hdiryellow3 <- "#FFBF00"
hdiryellow2 <- "#FFE187"
hdiryellow1 <- "#FFF9E6"

hdirblue2b <- "#048eb7"
hdirblue_f4 <- "#047FA4"

hdirfarge1 <- "#025169"
hdirfarge2 <- "#0069E8"
hdirfarge3 <- "#7C145C"
hdirfarge4 <- "#047FA4"
hdirfarge5 <- "#C68803"
hdirfarge6 <- "#38A389"
hdirfarge7 <- "#6996CE"
hdirfarge8 <- "#366558"
hdirfarge9 <- "#BF78DE"
hdirfarge10 <- "#767676"

colors_kjonn <- c(hdirfarge10, hdirfarge2, hdirfarge3)
show_col(colors_kjonn, ncol = 3)
```

https://designsystem.helsedir.no/0410c9e94/p/736ae4-farger
```{r}
hdirblue3 <- "#025169"
hdirblue2 <- "#047FA4"
hdirblue1 <- "#8CE2FD"
hdirblue1b <- "#9ce0f5"
hdirblue1c <- "#83cfe6"
fhiblue1b <- "#65ADC3"
hdirgreen3 <- "#366558"
hdirgreen2 <- "#95DBCA"
hdirgreen2b <- "#78d1bb"
hdirgreen_f6 <- "#38A389"
hdirgreen1b <- "#c5e9df"
hdirgreen1c <- "#dff7f0"
hdirgreen1d <- "#cff3e9"
hdirgreen1 <- "#EBF5F2"
hdirwhite <- "#FFFFFF"

colors_LFK <- c(hdirblue3, hdirblue2, hdirgreen2)
```

# Data prep


```{r}

map_levels(syke)

```

## Filtering
```{r}

syke_prep <- syke %>% 
  filter(ALDER == "20_66",
         AAR_end >= 2020,
         ) %>% 
  mutate(KJONN_text = case_when(KJONN == 1 ~ "Menn",
                                KJONN == 2 ~ "Kvinner",
                                KJONN == 0 ~ "Kjønn samlet",
                                ),
        )

```


```{r}
syke_plot_bydel <- left_join(jointable, syke_prep) %>% 
  filter(GEO_type %in% c("bydel", "kommune")) %>% 
  mutate(prikk = ifelse(is.na(MEIS), 3, NA_real_)) %>% 
  mutate(label_metadata = syke_data,
         label_kilde = "Kilde: SSB",
         ) %>% 
  mutate(plotid = ifelse(grepl("^301", plotid), paste0("0", plotid), plotid)) # ensure 4 digit Oslo geo code

map_levels(syke_plot_bydel)

syke_plot_bydel

```


# Linjeplot LFK

## testplot

```{r}
map_levels(syke_plot_bydel)
```

NO missing
```{r}
missing <- select_prikk(syke_plot_bydel %>% 
                          filter(AAR_end >= 2020), GEO_type = "bydel")

missing

# all missing
syke_plot_bydel %>% 
  filter(GEO_type == "bydel",
         AAR_end >= 2020,
         ) %>% 
  group_by(plotid) %>% 
  filter(all(is.na(MEIS))) %>% 
  pull(plotid) %>% 
  unique()

syke_plot_bydel %>% 
  filter(plotid %in% c(missing),
         GEO_type == "bydel",
         AAR_end >= 2020,
         ) %>%
  select(GEO, Sted) %>% 
  unique()
```

```{r}
min_max <- minmax(syke_plot_bydel)
min_max
```


# Linje KJONN

```{r fig.height=5, fig.width=7}
x_text_size <- 11.5
y_text_size <- 11.5
axis_title_size <- 14
legend_size <- 13.5
stripsize <- 14
colwidth <- 0.7
ymax <- 20 

misstext_bydel <- "Manglende eller utilstrekklig \ntallgrunnlag i bydelen"
#isstext_fylke <- "Kommunetall mangler, fylkestall vises"
colors_kjonn <- c(hdirfarge10, hdirfarge2, hdirfarge3)
ylab <- "Andel standardisert (prosent)"

theme_OVP_line <- theme_cowplot() +
    theme(legend.position = "bottom",
          legend.title = element_blank(),
          legend.text = element_text(size = legend_size),
          legend.margin=margin(t=-6),
          legend.key.width = unit(1.9, 'cm'),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size = x_text_size),
          axis.text.y = element_text(size = y_text_size),
          axis.title.y = element_text(size = axis_title_size),
          strip.background = element_blank(),
          #strip.text = element_blank(),
          line = element_blank(),
          axis.ticks.y = element_line(color = "grey50"),
          panel.border = element_rect(fill = NA, color = "grey50"),
          panel.grid.major = element_line(size = 0.2, color = "grey90"),
          panel.grid.minor.y = element_line(size = 0.1, color = "grey90"),
          plot.caption = element_text(hjust=c(0,1), size = c(5,10)),
          plot.margin = margin(t = 2, r = 5, b = 2, l = 5),
          panel.background = element_rect(color = "white"), 
          )

do_lineplot <- function(data, missing_bydelstall = FALSE, missing_plot_fylke = FALSE) {
  
  highest_plotorder <- data %>% filter(plotorder == max(plotorder)) %>% pull(Sted)

    p <- data %>% 
      ggplot(aes(x = AAR_end, y = MEIS,
                 color = reorder(KJONN_text, KJONN),
                 )) +
      geom_line(aes(group = reorder(KJONN_text, KJONN)),
                size = 1.5) +
      ggrepel::geom_label_repel(data = data %>% 
                                  filter(AAR_end == max(AAR_end)),
                                aes(x = AAR_end, y = MEIS,
                                    label = paste0(round(MEIS,1),"%")),
                                size = 4,
                                nudge_x = 0.5,
                                color = "black",
                                direction = "y",
                                force = 0.1,
                                box.padding = 0.1,
                                max.overlaps = Inf,
                                min.segment.length = 1,
                                ) +
      theme_cowplot() +
      scale_color_manual(values = c(colors_kjonn)) +
      geom_vline(xintercept = 0) +
      scale_y_continuous(limits = c(0, ymax),
                         breaks = seq(0,100,5),
                         expand = expansion(mult = c(0, 0)),
                         ) +  
      scale_x_continuous(limits = c(min_max$min_aar, min_max$max_aar+0.7),
                         breaks = seq(min_max$min_aar, min_max$max_aar, 1),
                         labels = abs) +
      coord_cartesian(clip = "off") +
      theme_OVP_line +
      facet_wrap(~ reorder(Sted, plotorder)) +
      #guides(color = FALSE) +
          labs(y = ylab,
           caption = c(syke_data, "Kilde: SSB"))

     # If missing values should be visualized, add these layers
  if (missing_bydelstall) {
    p <- p +
       geom_text(data = data %>% filter(Sted == highest_plotorder),
              aes(label = misstext_bydel,
                  #group = reorder(Sted, plotorder),
                  ),
                 x = (min_max$min_aar+min_max$max_aar)/2,
                 y = ymax-2,
                 size = 4,
                 check_overlap = TRUE,
                 inherit.aes = FALSE)
  }
  
  return(p)

  
}

syke_plot_bydel %>% 
  filter(GEO_type %in% c("bydel", "kommune"),
         plotid == "030101") %>% 
  do_lineplot(.)

#ggsave("plots/syke_KJONN_LK.png", height = 5, width = 7, bg = "white")

```


# MISSING

No all missing
```{r}

missing <- select_prikk(syke_plot_bydel, GEO_type = "bydel")

missing

# all missing
syke_plot_bydel %>% 
  filter(GEO_type == "bydel") %>% 
  group_by(plotid) %>% 
  filter(all(is.na(MEIS))) %>% 
  pull(plotid) %>% 
  unique()

syke_plot_bydel %>% 
  filter(plotid %in% c(missing),
         GEO_type == "bydel") %>%
  select(GEO, Sted) %>% 
  unique()
```


# MULTIPLOT

```{r}

dirname <- "Figur4_BYDEL"
figposnr <- "5"

res_dir <- paste0("plots/", dirname)
dir.create(res_dir, recursive = TRUE)

# Plot Level 1 - no missing
plots_1 <- syke_plot_bydel %>% 
  filter(GEO_type %in% c("bydel", "kommune")) %>% 
  filter(!plotid %in% missing) %>%  # Filter plotid with no missing
  split(.$plotid) %>% 
  map(~do_lineplot(.x))
  #map(~do_plot(.x, include_missing = FALSE))

# Plot Level 2 - missing
plots_2 <- syke_plot_bydel %>%
  filter(GEO_type %in% c("bydel", "kommune"),
         plotid %in% missing) %>%  # Filter plotid with no total missing
  split(.$plotid) %>%
  map(~do_lineplot(.x, missing_bydelstall = TRUE))

# Combine both lists of plots
all_plots <- c(plots_1, plots_2)

# Generate file paths using `outdir` and `figposnr`
paths <- stringr::str_c(names(all_plots), "_", figposnr, "_tema", ".png")

# Save plots
pwalk(list(paths, all_plots), ~ggsave(filename = .x, plot = .y, 
                                      height = 5, width = 7, dpi = 300, bg = "white", 
                                      path = res_dir))

length(plots_1)+length(plots_2)
length(all_plots)
```







